Trending assessment for nowcasts, measurements, and forecasts

### 2024-03-29 paper.tex
Move everything to one file

\title{Trending assessment for nowcasts, measurements, and forecasts}
%\author{Oliver Grothe, Bolin Liu, Jonas Rieger}
\author[1]{Oliver Grothe}
\equalcont{These authors contributed equally to this work.}
\author[1]{Bolin Liu}
\equalcont{These authors contributed equally to this work.}
\author*[1]{Jonas Rieger}
\equalcont{These authors contributed equally to this work.}

@@

\section{Introduction}\label{sec:introduction}
\input{sections/01-introduction}

%\section{Application areas and Notation} \label{sec:notation}
%\input{sections/02-notation}

\section{Trending assessment}\label{sec:trending}
\input{sections/03-trending}

%\section{Simulation Studies (?)}
%
%Eher nicht, weil angewandtes Paper

\section{Application to measurement, nowcasting, and forecasting data} \label{sec:application}
\input{sections/04-application}

>>>>

\title{Trending assessment for nowcasts, measurements, and forecasts}
%\author{Oliver Grothe, Bolin Liu, Jonas Rieger}
\author[1]{Oliver Grothe}
\author[1]{Bolin Liu}
\author*[1]{Jonas Rieger}

@@

\section{Introduction}\label{sec:introduction}
The evaluation of measurement, prediction, and forecasting methods becomes increasingly important and sophisticated as technologies and data availability enable their application in more and more fields.
Conventionally, differences between predictions (or measurements) and target values, such as the \ac{rmse} or \ac{mae}, form the basis of method evaluation.
These measures do not consider trending, that is, whether the right direction of change is predicted or measured.
However, the information about an increase or decrease is crucial to detect changes in the underlying process early.
In the following, we provide a more specific overview of the characteristics and current evaluation schemes of the three application fields, forecasting, nowcasting, and measurement, and highlight why trending evaluation is highly relevant in the respective fields.

Forecasting methods predict the future based on historical data, patterns, and exogenous factors.
In medicine, forecasting is used in various applications, such as population and personal health~\parencite[see, e.g., the review in]{Soyiri2013}, emergency department demand forecasting~\parencite{Jones2008,Rostami-Tabar2023}, and emergency medical services~\parencite{HaugsboHermansen2021}.
The forecast is computed based on the current value of the quantity of interest and an estimate of the development until the target time.
In the current practice, a forecasting method is usually evaluated in terms of its prediction accuracy, measuring the deviation of the prediction from the actual outcome of the target variable.
Popular measures are scale-dependent measures such as the \ac{rmse}, measures based on percentage errors such as the root mean squared percentage error, or probabilistic scoring rules~\citep[see the review in][]{hyndman2006another}.
They are not capable of assessing the trending ability discussed above.
In a prediction competition on armed conflicts, trending recently gained attention as \citet{Vesco2022} used the novel \ac{tadda} score with a trending component for evaluation.
However, the score poses an unintuitive incentive to forecasters and is thus theoretically problematic~\parencite{Bracher2023}.
This paper focuses on pure trending evaluation that accompanies traditional scores.

Methodologically evolved from forecasting; nowcasting methods focus on predictions for the present, the immediate future, and the recent past \citep{banbura2013now}.
Nowcasting has its origins in meteorology, and the methods were initially developed to describe the current state of the weather in detail and to predict the expected change on a time scale of a few hours~\citep{browning1989nowcasting,schmid2019nowcasting}.
%In economics, nowcasting is used to predict statistics on the current economic situation, for example, the gross domestic product, which is collected with low frequency and is available with a considerable time delay \citep{banbura2013now}.
In medicine, epidemic nowcasting, for example, can assess the current situation during an ongoing epidemic, considering the main pathogenic, epidemiological, clinical, and socio-behavioral factors~\citep{wu2021nowcasting} or correct the daily COVID case numbers for events that have occurred but have not yet been reported~\citep{gunther2021nowcasting}.
Nowcasting methods use high-frequency indicators related to the target variable and estimate the value of a target variable for a specific time based on current preliminary measurements, which are finalized with a considerable time delay.
Thus, the nowcasts can produce early and ongoing estimates for the target variable during the relevant period~\citep{castle2017forecasting}.
Like forecasts, nowcasts are often evaluated and compared in the literature based on performance measures such as the \ac{rmse}~\citep{gunther2021nowcasting} or probabilistic scoring rules~\citep{Wolffram2023}.
The aspect of trending ability is not considered in the literature to the best of our knowledge.
However, trending evaluation adds valuable information on the methods' capability of predicting the development of the target variable, for example, the number of cases of an epidemic, where increases can be an early sign of the beginning of a wave.
The epidemic's development, in turn, can be the foundation of decisions on introducing or canceling policy measures.

Measurement is more technical and aims to obtain accurate and reliable data about the current state of a system based on sensors.
A parameter or variable can be measured regularly over a certain period to evaluate the system's development.
When introducing new, cheaper, simpler, or less invasive measurement methods, they are evaluated against current measurement techniques, called the gold standard, by measuring the same quantity in different settings, such as various individuals or environments.
Various indices, such as the interclass correlation coefficient or Lin's concordance correlation, have been proposed to check the reproducibility of the measurement or to compare different measurement methods~\citep{lawrence1989concordance,koo2016guideline} in addition to paired t-tests to determine systematic differences between two measurement series~\citep{watson2010method}.
Graphical tools, such as the Bland-Altman diagram, are used to visualize the differences between the methods in relation to their mean value \citep{bland1986statistical}.
Trending considerations, that is, the consistency of the test method's development with the gold standard, is a field of active research in the last years \citep{Saugel2015,saugel2018error,hiraishi2021concordance}.

The main contributions of this paper are manifold.
We formalize trending and present different variants of trending measures that consider either noiseless data or data with noise and small non-informative changes.
We introduce the conditional trending plot, a new graphical method for assessing local trending behavior, and review bootstrap methods for calculating confidence intervals.
We extend the concept of trending to probabilistic predictions.
The methods are applied to three applications from the abovementioned fields: COVID-19-nowcasting, emergency department arrival forecasting, and blood pressure measurement.
The ready-to-use code for trending evaluation from the applications is available on ... .
The trending evaluation methods are intended to complement the current evaluation and reveal additional information on the methods considered.

The remainder of this paper is structured as follows.
Section~\ref{sec:trending} formalizes trending and investigates several extensions, such as noise-aware methods, confidence intervals, and probabilistic forecasts and nowcasts.
We introduce a new graphical method for evaluating trending, the conditional trending plot, which can simultaneously assess different intervals and asymmetries.
In Section~\ref{sec:application}, we show the results of applying our method to several practical examples from measurement, forecasting, and nowcasting.
Hence, we provide blueprints for applying trending evaluation in practice and make remarks on interpreting the measures.
We conclude the paper in Section~\ref{sec:conclusion}.

\section{Trending assessment}\label{sec:trending}

As discussed in Section~\ref{sec:introduction}, the term \enquote{trending} denotes the alignment of a detected true development with the development indicated by a measurement, a nowcast, or a forecast.
In this section, we want to specify trending and define different methods of measuring and visualizing trending.
In general, trending can be assessed for different horizons $l$ separately, and the considered horizons' relevance must be determined based on the task.
Throughout the section, we call a measurement, nowcast, or forecast a \textit{prediction} or \textit{signal} whenever the application area is irrelevant.
Before introducing measures to evaluate trending, we remark on calculating the predicted change and notation in Section~\ref{subsec:notation}.
Section~\ref{subsec:trending-four-quadrant-plot} introduces trending basics and visualizes trending in a synthetic example.
In Section~\ref{subsec:trending-measures}, various methods from other disciplines and tasks are applied and examined.
The approaches are extended to account for noise and non-systematic effects in Section~\ref{subsec:trending-noise}, and bootstrap methods for computing confidence intervals for the measures in the previous sections are briefly introduced.
Section~\ref{subsec:trending-cond-prob} introduces a new graphical method for local trending assessment.
Trending assessment for probabilistic nowcasts and forecasts is treated in Section~\ref{subsec:trending-probabilistic}.

\subsection{Computing changes and notation}\label{subsec:notation}

As noted above, trending refers to observed and predicted changes over a time horizon $l$.
The \textit{observed} change is straightforward to compute for all types of signals.
Let $\mathbf{y} = (y_t)_{t=0}^T$ denote the true values for nowcasting or forecasting, or gold standard measurements up to time $T$.
The sequence of observed changes is then given by the differences of values in $\mathbf{y}$ with horizon $l$, that is,
\begin{equation}\label{eq:diffy}
    \diffylt = (y_{t} - y_{t-l}) \quad \text{for}\ t = l, \dots, T.
\end{equation}

\begin{table}
    \centering
    \scriptsize
    \begin{tabular}{l l}
        \toprule
        Application & Predicted change computation \\
        \midrule
        Measurement & $(x_{t} - x_{t-l})_{t=l}^T $\\
        Nowcasting & $
\diffxl =
\begin{cases}
(x_{t|t} - x_{t-l|t})^T_{t=l}, & \text{if } y_{t-l} \text{ is not known at time } t, \\
(x_{t|t} - y_{t-l})^T_{t=l}, & \text{otherwise}.
\end{cases}
$ \\
        Forecasting & $ \diffxl = (x_{t|t-l} - y_{t-l})^T_{t=l}$\\
        \bottomrule
    \end{tabular}
    \caption{Computation of the predicted change in the different applications. For nowcasting and forecasting, signals $x_{t | \tau}$ refer to values issued at $\tau$ with a target time $t$. For measurement, $x_t$ denotes the test device measurement at time $t$.}
    \label{tab:notation}
\end{table}

The definition of the \textit{predicted} change depends on the context;
Table~\ref{tab:notation} summarizes the notation for the different signal types and the computation of the predicted change.
For nowcasting, let $x_{t | \tau}$ denote the nowcast for time $t$ computed with the knowledge of time $\tau$.
We call $t$ the \textit{forecast time} and $\tau$ the \textit{issue time}.
The predicted change is computed by
\begin{equation}
  \diffxl =
\begin{cases}
(x_{t|t} - x_{t-l|t})^T_{t=l} & \text{if } y_{t-l} \text{ is not known at time } t, \\
(x_{t|t} - y_{t-l})^T_{t=l} & \text{otherwise}.
\end{cases} \label{eq:diffxl_nowcasting}
\end{equation}
When computing the predicted change of a nowcast for a time $t$, we use the best knowledge available at that time $t$, and the true value might not be known yet.
If the true value $y_{t-l}$ is known at time $t$, the predicted change is computed by the difference between the nowcast and the true value, as $y_{t-l}$ is also known by the nowcaster and incorporated into the nowcast.
Through the computation in Equation~\eqref{eq:diffxl_nowcasting}, the predicted change can be computed with the knowledge of the nowcaster at time $t$.

The notation is similar for forecasting: Let $x_{t | \tau}$ denote the forecast for forecast time $t$ and issue time $\tau$.
The predicted change is computed by
\begin{equation} \diffxl = (x_{t|t-l} - y_{t-l})^T_{t=l}\label{eq:diffxl_forecasting}
\end{equation}
with the same structure as in the nowcasting case.
The indices are chosen to be consistent with $\diffyl$.
If the true value $y_{t-l}$ is not known at time $t-l$, a similar modification as in Equation~\eqref{eq:diffxl_nowcasting} can be made.

The distinction between forecast and issue time is unnecessary in measurement analysis, as the measurement is typically available with a very short time lag.
Thus, $x_t$ denotes the test method measurement for time $t$.
The computation
\begin{equation}
  (x_{t} - x_{t-l})_{t=l}^T \label{eq:diffxl_measure}
\end{equation}
yields the change by the test method.
It is computed purely by the test method without the gold standard $y_t$ to analyze whether the gold standard and test method changes are consistent.
Accordingly, $y_{t-l}$ is not used in the computation even if known at time $t$ in contrast to forecasting and nowcasting.

In applications, signal data are often not available for all time steps, for example, due to technical problems or delays in data transfer (see the examples in Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}).
We refer to time steps for which either signal or true values, or both, are unavailable as missing values.
Systematical missing values could lead to a biased estimate of the trending ratio, and missing data should be inspected for any underlying patterns.
If the missing values are not systematic, random, and occur scarcely, data pairs with missing values can be excluded from the data to calculate the measures~\parencite[see][Section 1.3]{VanBuuren2018}.
Note that in the case of measurement data, one missing value in the time series leads to two undefined differences in the change series; that is if $x_t$ is missing, $\diffxt$ and $\diffxt[t+l]$ are undefined; if an observation $y_t$ is missing, $\diffyt$ and $\diffyt[t+l]$ are undefined.
The data pair is excluded even if the corresponding nowcast or forecast is available.

\subsection{Basics of trending and four-quadrant plots}\label{subsec:trending-four-quadrant-plot}

Formally, the trending assessment is the same for all three application areas, given the notation for the respective application of Section~\ref{subsec:notation}.
In the following, we omit the difference horizon $l$ for ease of notation; $\diffx$ and $\diffy$ refer to $\diffxl$ and $\diffyl$ for a common horizon $l$.
As noted above, a signal's trending is perfect if all the change directions are predicted correctly; that is, the sign of all elements of $\diffx$ and $\diffy$ coincide.
Consequently, when evaluating the trending, we examine the statistical consistency of $\sign(\diffx)$ and $\sign(\diffy)$.
A simple yet insightful method is the four-quadrant plot~\citep[see, e.g., ][]{perrino1998intraoperative,Saugel2015}.
In a four-quadrant plot, the occured changes are plotted with the predicted changes, that is, $(\diffyt, \diffxt)$ for $t = l, \dots, T$.
Thus, the x-axis of a four-quadrant plot shows the true value differences, whereas the y-axis displays the prediction data differences.
Points in the green upper right and lower left quadrants reflect a correct trending for the respective time step, whereas points in the remaining red quadrants show incorrect predicted changes.

Figure~\ref{fig:trending_basic_4q} displays a basic four-quadrant plot.
Points 2, 3, 5, and 6 show trending, whereas points 1, 4, and 7 count for anti-trending behavior.
Figure~\ref{fig:trending_basic_4q_sample} shows a four-quadrant graph for a more extensive simulated data set with $T=1461$, for example, four years of daily data.
Data generation is described in Appendix~\ref{subsec:app-trending-data-generation}.
The four-quadrant plot resembles a butterfly-like shape, with more data in the upper right and lower left quadrants than in the remaining quadrants.

The four-quadrant plot is intuitive to interpret, and the magnitude and direction of change are shown simultaneously.
It can be extended by including information on the time index in the point color to reveal effects over time.
In Figure~\ref{fig:trending_basic_4q_sample_color}, the point colors turn from blue to green for higher time indices $t$, that is, more recent values; $(\diffyt[l], \diffxt[l])$ is blue and turns green until $(\diffyt[T], \diffxt[T])$.
However, four-quadrant plots become crowded for larger datasets, and sequential information on the differences is complex to assess thoroughly.

There are other visualization techniques, such as polar plots or the Bland-Altmann analysis.
However, they lack the four-quadrant plot's clarity and intuition without adding more information on trending~\citep{Saugel2015}.

\begin{figure}
\centering
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_without_excl}
\caption{Basic four-quadrant plot.} \label{fig:trending_basic_4q}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_box}
\caption{Four-quadrant plot with rectangular exclusion area.}\label{fig:trending_basic_4q_excl_box}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_axis}
\caption{Four-quadrant plot with horizontal exclusion area.} \label{fig:trending_basic_4q_excl_axis}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_cross}
\caption{Four-quadrant plot with cross-shaped exclusion area.}\label{fig:trending_basic_4q_excl_cross}
\end{subfigure}%
\caption{Illustrations of the four-quadrant plot with sample points and with and without exclusion areas. }
\label{fig:trending_4q}
\end{figure}

\subsection{Trending ratio and other measures}\label{subsec:trending-measures}

Analyzing the number of points in the green versus red quadrants is a standard approach in the trending assessment of measurement data \citep{Critchley2010,Saugel2015}.
With that, we mainly estimate the probability of trending $P(\diffxrv \diffyrv > 0)$, where $\diffyrv$ and $\diffxrv$ denote random variables for future incremental changes.
Since $z_1 z_2 > 0$ imposes the same condition as $\sign(z_1) = \sign(z_2)$ ($z_1, z_2 \in \R \setminus \{ 0 \}$), the standard estimator for $P(\diffxrv \diffyrv > 0)$ is
\begin{equation}
    \acc (\diffx, \diffy) \coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0}}{\card{T}}.\label{eq:acc}
\end{equation}
Here, the numerator counts the number of same-sign-changes, while the denominator is the number of considered pairs $(\diffyt, \diffxt)$.
Thus, $\acc$ is the proportion of concordant changes on all changes.
We refer to this estimator as the \textit{trending ratio} of the prediction and set $\mathcal{T} = \{l, \dots, T\}$.
Visually, the measure computes the fraction of points in the upper right or lower left quadrant.
Similar evaluations are used in other scientific areas, for example, with contingency tables in dichotomous forecasting or with confusion matrix in classification analysis~\citetext{see, e.g., the introductions in \citealt{James2021}, Ch. 4, and \citealp{Jolliffe2012}, Ch. 3}.
Many other measures can be adapted from those fields to deepen the trending analysis.
Two simple measures that focus on either a positive or negative predicted change are the positive and negative trending ratios $\accp$ and $\accm$, respectively.
They are defined as
\begin{align}
    \accp (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0}},\ \text{and} \label{eq:accp}\\
    \accm (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0}}.\label{eq:accm}
\end{align}
They estimate the probability of a correct prediction of the direction of change, given that the signal direction is positive or negative, that is, $P(\diffxrv \diffyrv > 0 | \diffxrv > 0)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0)$.

Rolling estimates of the above measures detect changes in performance over time and can give a sharper estimate of the current trending ability.
For the trending ratio, a rolling estimate with a backward-looking window of length $w$ at time $t$ is given by
\begin{equation*}
    \acc_{t; w} (\diffx, \diffy) \coloneqq \frac{\sum_{t^\star = t-w + 1}^{t} \ind{\diffxt[t^\star] \diffyt[t^\star] > 0}}{w}.\label{eq:acc_rolling}
\end{equation*}
Backward-looking windows estimate the trending ratio at a time $t$ considering the $w$ time steps before time $t$.
The window length $w$ controls the smoothing of the estimate; a larger $w$ gives smoother results, while a small $w$ focuses on local variations.
Plotting the rolling estimates for $t = w-1, \dots, T$ yields an estimate of the trending ability over time.
Figure~\ref{fig:trending_ratio_time_series} depicts a rolling window estimate of the trending ratio for the simulated data of Figures~\ref{fig:trending_basic_4q_sample} and~\ref{fig:trending_basic_4q_sample_color}.
Thus, the yearly course of the prediction trending ratio can be detected.
The trending ability of the signal has a strong sinus-shaped seasonality with a trending ability peak after a quarter of a year and a low point after three quarters.
While colored four-quadrant plots, as in Figure~\ref{fig:trending_basic_4q_sample_color}, illustrate ongoing overall drifts in the trending ability, seasonal aspects are only revealed in rolling window estimates.

\begin{figure}
    \centering
    \begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_sample_without_time}
\caption{Four-quadrant plot with simulated data.}\label{fig:trending_basic_4q_sample}
\end{subfigure}\hspace{0.01\textwidth}
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_sample_with_time}
\caption{The same data is colored according to the time index $t$, the greener, the later.}\label{fig:trending_basic_4q_sample_color}
\end{subfigure}\hspace{0.01\textwidth}
\begin{subfigure}[t]{.48\textwidth}
    \includegraphics{plots/illustrative_examples/trending_ratio_time_series.pdf}
    \caption{Rolling estimate of the trending ratio over time with window length 91. }\label{fig:trending_ratio_time_series}
    \end{subfigure}%
    \caption{Visualizations for data with a time-varying trending ratio. We defer information on the data generation process to the appendix (see \ref{subsec:app-trending-data-generation}). The trending ratio for the entire data set is $\mu = 0.7577$. The strong seasonality of the trending ratio becomes visible in Figure~\ref{fig:trending_ratio_time_series}. The green curve $k_t$ shows the theoretical probability that $\diffxt$ has the same sign as $\diffyt$ for each time step. The rolling estimates are shifted to the right compared to $k_t$ as the windows look backward.}
\end{figure}

\subsection{Accounting for noise and non-informative small changes and bootstrapping confidence intervals}\label{subsec:trending-noise}

The above measures can be extended to account for information on the point's location within the quadrant.
For example, points close to the zero point may have less explanatory power or may be less reliable than points far away from zero on one of the diagonals.
Suppose noise or non-systematic effects are present in the true values or predictions.
In that case, a point's assignment to a quadrant can be driven by noise instead of a systematic trending ability of the signal.
This is more likely for points with at least one small coordinate.

Using an exclusion area around the zero point, as further defined below, is a straightforward and highly interpretable extension of the measures of Section~\ref{subsec:trending-measures} accounting for such effects\citep[see, e.g.,][]{Saugel2015,Critchley2010}.
Points within that area are omitted in the four-quadrant plot and the calculation of the measures.
In particular, the prediction models are likely to have a noise component; thus, $\diffx$ should be subject to an exclusion area.
We denote the measures of Equations~\eqref{eq:acc},~\eqref{eq:accp} and~\eqref{eq:accm} accounting for an exclusion area $E$ by
\begin{align}
    \acceps (\diffx, \diffy, E) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffx \diffy > 0} \ind{(\diffyt, \diffxt) \notin E}}{\sum_{t \in \mathcal{T}} \ind{(\diffyt, \diffxt) \notin E}},\label{eq:acceps}\\
    \accpeps (\diffx, \diffy, E) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0, (\diffyt, \diffxt) \notin E}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0, (\diffyt, \diffxt) \notin E}}, \ \text{and} \label{eq:accpeps}\\
    \accmeps (\diffx, \diffy, E) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0, (\diffyt, \diffxt) \notin E}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0, (\diffyt, \diffxt) \notin E}}. \label{eq:accmeps}
\end{align}
The additional argument $(\diffyt, \diffxt) \notin E$ in the indicator ensures that only points not lying in the exclusion are evaluated when computing the measures.
The measures are then estimators for the probability of trending, given that the point's location is not driven by noise or non-informative changes.

The estimators accept various shapes of the exclusion area.
Figure~\ref{fig:trending_4q} visualizes examples of exclusion area shapes.
A rectangular exclusion area, $E = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \land (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ ($\varepsilon_x, \varepsilon_y > 0$), leaves out points that are small in both components and, therefore, likely to be driven by noise.
Points are included if the true value or signal is unlikely to be noise-driven.
In the example graph in Figure~\ref{fig:trending_basic_4q_excl_box}, only point 1 is excluded.
An exclusion area along one axis, for example, $E = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x)\}$ for $\varepsilon_x > 0$, removes points in which one of the components could change sign by a small amount of noise.
This particularly suits signals where small amounts of noise are inevitable.
A cross-shaped exclusion area, $E = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \lor (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, along both axes accounts for the sign reversal in both components.
A point is only included if no component is likely to be driven by noise.
For the latter two methods, points 1 and 7 in Figure~\ref{fig:trending_basic_4q_excl_axis} or 1, 6, and 7 are excluded in Figure~\ref{fig:trending_basic_4q_excl_cross}, respectively.

In most applications, the shape and size of the exclusion area can be chosen based on domain knowledge or expert opinions.
The size determination can also be based on a proportion of the total variance or the total range of the data; for example, the 10\% smallest absolute values in each component determine the exclusion area size.
A third approach is to visualize the trending ratio for different sizes of $E$ and thus inspect the effects of the exclusion area size on the estimates.
For examples of such plots, see Section~\ref{sec:application-covid}.

Confidence intervals can account for the estimation uncertainty of the measures above.
Bootstrap confidence intervals are based on resampling and not on parametric assumptions as classical confidence intervals are~\parencite[for introductions see][]{Hesterberg2011,Bittmann2021}.
Many new samples are drawn with replacement from the dataset, and the statistic of interest is computed for each sample, yielding an estimate for the distribution of the statistic of interest.
Based on the derived \enquote{new} samples of the statistic, the confidence intervals can be derived through differently sophisticated bootstrapping methods.
We use the \ac{bca} approach for bootstrapping in the following, as it holds the confidence level for small and large samples and has a moderate computation time (see the simulation study in Appendix~\ref{subsec:appendix-trending-bootstrap}).

\subsection{The conditional trending plot}\label{subsec:trending-cond-prob}

The estimators described above give information on the probabilities $P(\diffxrv \diffyrv > 0 | \diffxrv \diffyrv \notin E)$, $P(\diffxrv \diffyrv > 0 | \diffxrv > 0, \diffxrv \diffyrv \notin E)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0, \diffxrv \diffyrv \notin E)$.
A still finer analysis might be gained by considering the conditional distribution $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ to assess the trending ability of a prediction for a certain change size $\diffxrv = \xcond$ of the signal.
Thereby, $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ denotes the probability of a correct trending given a predicted change of $\xcond$.
Thus, if a change of $\xcond$ is observed in practice, one can directly assess the credibility of this specific signal.
A multivariate \acf{kde} facilitates a continuous estimation of $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ by estimating the components $f_{\diffxrv, \diffyrv}$ and $f_{\diffxrv}$ of
\begin{align*}
P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond) = \begin{cases}
                                              \int_{-\infty}^0 \frac{f_{\diffxrv, \diffyrv}(\xcond, y)}{f_{\diffxrv}(\xcond)} \ \textrm{d} \: y & \text{if } \xcond < 0, \\
                                              \int_{0}^{\infty} \frac{f_{\diffxrv, \diffyrv}(\xcond, y)}{f_{\diffxrv}(\xcond)} \ \textrm{d} \: y & \text{if } \xcond > 0, \\
\end{cases}
\end{align*}
for $\xcond \neq 0$ through a \ac{kde}.
A comprehensive introduction to multivariate \ac{kde} can be found in \textcite{Gramacki2018}, and implementations are available in many programming languages, for example, in the \verb|statsmodels| in Python~\parencite{Seabold2010}.
The \ac{kde} yields estimates for $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ for all values of $\xcond \in \R$.
Multivariate \ac{kde} takes a kernel and bandwidth selector as modeling parameters.
We advise using a Gaussian kernel and the cross-validation maximum likelihood as bandwidth selector (see Appendix~\ref{subsec:appendix-kde}).

Assessing $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ graphically by drawing $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ against $\xcond$ eases the simultaneous evaluation of various $\xcond$.
Furthermore, the graph facilitates the comparison of various methods in a single graph, and asymmetries of $P(\diffxrv \diffyrv > 0 | \diffx = \xcond)$ with respect to $x$ in the trending ability can be detected.
We refer to the plot as a \textit{conditional trending plot}.

\subsection{Probabilistic Evaluation}\label{subsec:trending-probabilistic}

In nowcasting and forecasting, probabilistic predictions have become more prevalent in recent years (see, for example, Sections~\ref{sec:application-covid} and~\ref{sec:application-eda} or the review in~\cite{Gneiting2014}).
In this section, we develop trending measures for probabilistic signals.
Probabilistic predictions issue a probability distribution for the quantity of interest based on their available information and, thus, include a point estimate and information on the prediction uncertainty and quantiles simultaneously.
For trending analysis, probabilistic predictions thus also contain a probability of a positive or negative change.
In this section, we extend the trending assessment to probabilistic predictions.

Probabilistic predictions can take the form of a \ac{cdf}, \ac{pdf}, or quantiles.
The \ac{cdf} is the most general and can be used to derive the others, given they exist.
Let us first assume that the prediction is a \ac{cdf}, and that $y_{t-l}$ is known at time $t$ (see Table~\ref{tab:notation}).
In Appendix~\ref{subsec:app-probabilistic-trending-evaluation}, we extend the analysis for predictions given as quantiles or unknown true values.

Let for a forecast $F_{t | t-l} (x)$ denote the predicted \ac{cdf} for forecast time $t$ and issue time $t - l$, where the index is analogous to the point prediction notation of Section~\ref{subsec:notation}.
Thereby, $F_{t | t-l} (x)$ yields the probability that the quantity of interest is at most $x$.
For trending assessment, we compare the predicted probability of positive change, denoted by $p_t$, with the occurrence of positive changes.
A positive change occurs for any value at $t$ larger than the true value $y_{t-l}$, as then the change compared to $y_{t-l}$ is positive.
The \ac{cdf} $F_{t | t-l} (y_{t-l})$ yields the predicted probability of any value at most $y_{t-l}$, a negative change.
Thus, $1 - F_{t | t-l} (y_{t-l})$ is the probability of a positive change.
As for the differences in Section~\ref{subsec:notation}, the computation differs slightly for nowcasts and forecasts.
For nowcasts, the probability is calculated by
\begin{equation*}
    p_t = 1 - F_{t | t-l} (y_{t-l})\quad t = l, \dots, T;
\end{equation*}
while for nowcasts it is given by
\begin{equation*}
    p_t = 1 - F_{t | t} (y_{t-l})\quad t = l, \dots, T,
\end{equation*}
with analogous derivations as above.
Let $z_t$ denote the indicator that the observed change at time $t$ is positive, that is,
\begin{equation*}
    z_t = \ind{\diffyt > 0} \quad t = l, \dots, T.
\end{equation*}
The probabilistic trending evaluation is then based on the probabilities $\mathbf{p} = (p_t)_{t=l}^{T}$ and the observed changes $\mathbf{z} = (z_t)_{t=l}^{T}$.
The predictive power of $\mathbf{p}$ for $\mathbf{z}$ can be assessed using probabilistic dichotomous forecast evaluation methods.
Dichotomous forecasts predict a binary outcome, such as a positive or negative change.
Numerical measures for forecast quality are scoring rules.

The \ac{bs} is a widely used scoring rule for dichotomous probabilistic forecasts~\citep{Brier1950}.
In our context, it is
\begin{equation*}
    BS (\mathbf{p}, \mathbf{z}) = \frac{1}{T-l+1} \sum_{t=l}^{T} (p_t - z_t)^2.
\end{equation*}
Thereby, lower values indicate a higher probabilistic trending ability of the considered method.
The \ac{bs} assesses the calibration and sharpness of the forecast and the observation simultaneously~\citep{Ranjan2010,Mitchell2011}.
Calibration refers to the statistical consistency of forecasts and observations; that is, the event occurs with the issued probability.
Sharpness refers to the spread of the forecast; higher probabilities are preferable as they convey a higher certainty.

Graphical methods are a standard tool for evaluating the calibration of probabilistic forecasts in detail.
In dichotomous forecasting, the reliability diagram is frequently used~\citep{Ranjan2010}.
The reliability diagram plots the observed frequency of the positive outcome against the (binned) predicted probability.
For example, it shows the proportion of observed increases, given the predicted probability of increase was approximately $0.7$.
Ideally, the predicted probability equals the observed frequency, and the reliability diagram is a 45-degree line.
Local deviations from the 45-degree line indicate a miscalibration of the forecast for specific probabilities.
Thus, the reliability visualizes the local and overall calibration simultaneously.
For an example of a reliability diagram, see Section~\ref{sec:application-eda}.

\section{Application to measurement, nowcasting, and forecasting data} \label{sec:application}
In this section, we evaluate the trending in three different applications.
In the first application in Section~\ref{sec:application-covid}, we consider the nowcasting of the seven-day hospitalization rate for COVID-19 in Germany.
Section~\ref{sec:application-eda} applies the trending assessment to forecast methods for the number of arrivals in a large emergency department.
In the last application in Section~\ref{sec:application_measurement}, we consider the trending assessment of non-invasive blood pressure measurements compared to invasive blood pressure measurements.

The examples illustrate the use of the methods of Section~\ref{sec:trending} and yield further insights into the measures and plots.
All code used to generate the analysis is available online.

\subsection{Covid nowcasting} \label{sec:application-covid}

Amid the COVID-19 pandemic, the importance of accurate and prompt nowcasts of the pandemic's progression has become evident.
Various indicators measured the pandemic's spread and severity.
In Germany, the seven-day hospitalization rate was established as a central steering measure for COVID-19 measures in November 2021, and the imposition of severe public restrictions was based on it~\citep{RobertKochInstitute2021}.
The Robert Koch Institute (RKI) provided preliminary daily reports on the seven-day hospitalization rate.
However, these reports were subject to severe delays and revisions in two sources.
The first source is technical delays in the reporting process, for example, due to different authorities passing the data to the RKI~\citep{RobertKochInstitute2024}.
The second, more systematic source is the structure of the seven-day hospitalization rate.
To a given date, all the cases are allocated whose first positive test result was on that date and who were hospitalized in relation to the disease in the following.
The seven-day hospitalization rate is the average number of those cases per $100,000$ inhabitants on the given date and six days before.
Thus, the final seven-day hospitalization rate can only be reported with a significant delay of more than 70 days, as the hospitalization of infected inhabitants can occur much later than the first positive test.
Nevertheless, as the seven-day hospitalization rate was considered a major indicator of the pandemic's development, many organizations and institutions started to issue nowcasts, including research teams and newspapers.
To collect nowcasts of the seven-day hospitalization rate by different nowcast groups, the COVID19-Nowcasting-Hub was established~\citep{ChairOfEconometricsAndStatisticsAtKarlsruheInstituteOfTechnology2024}.
The nowcasts contain the seven-day hospitalization rate's predictive mean, median, and other quantiles.
